---
title: "AI-Assisted Pipeline for Ants"
author: "Daniel Flo"
date: "`r format(Sys.Date(), '%d-%m-%Y')`"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
encoding: UTF-8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# Fix encoding for PaperQA2 HTML file
html_file <- "C:/Users/dafl/OneDrive - Folkehelseinstituttet/VKM Data/27.02.2025_maur_forprosjekt_biologisk_mangfold/data/species/1317433/1317433_PaperQA2_20251125_145755.html"

if (file.exists(html_file)) {
  content <- readLines(html_file, encoding = "latin1", warn = FALSE)
  writeLines(content, html_file, useBytes = FALSE)
}
```


## What is RAG

Retrieval-Augmented Generation (RAG) combines large language models with targeted document retrieval to extract information from specific sources rather than relying on pre-trained knowledge.

**Key advantages:**

- Grounded in source evidence
- Reduced hallucinations
- Access to specialized domain literature

## RAG workflow components

- **Query:** Species-specific interaction questions
- **Retrieval:** PDF chunks with relevant mentions
- **Augmentation:** LLM receives context from retrieved chunks
- **Extraction:** Structured interaction data with source validation


## PaperQA2 RAG output example: Anoplolepis gracilipes
```{r paperqa-report, results='asis'}
paperqa_file <- "C:/Users/dafl/OneDrive - Folkehelseinstituttet/VKM Data/27.02.2025_maur_forprosjekt_biologisk_mangfold/data/species/1317433/1317433_PaperQA2_20251125_145755.html"

if (file.exists(paperqa_file)) {
  content <- readLines(paperqa_file, encoding = "UTF-8", warn = FALSE)
  # Extract body content only
  body_start <- grep("<body", content, ignore.case = TRUE)[1]
  body_end <- grep("</body>", content, ignore.case = TRUE)[1]
  if (!is.na(body_start) && !is.na(body_end)) {
    cat('<div style="height: 400px; overflow: auto; border: 1px solid #ccc; padding: 10px;">')
    cat(content[body_start:body_end], sep = "\n")
    cat('</div>')
  }
} else {
  cat("PaperQA2 report not found.")
}
```


## Step 3: Complete Report using GPT-Researcher

- Deep research (GPT-Researcher) on known interactions
- Focus: establishment, climate, impact
- Produce 1-2 pages with references

More information: [github.com/assafelovic/gpt-researcher](https://github.com/assafelovic/gpt-researcher)

## GPT-Researcher workflow
```{r gpt-flow-diagram, out.width="70%", fig.align="center"}
flow_image <- file.path(
  "C:/Users/dafl/OneDrive - Folkehelseinstituttet/VKM Data",
  "04.09.2025_Betula_Nordic_Network/gpt_flow.png"
)

if (file.exists(flow_image)) {
  knitr::include_graphics(flow_image)
}
```

## GPT-Researcher example: Anoplolepis gracilipes
```{r gpt-researcher-report}
gpt_file <- file.path(
  "C:/Users/dafl/OneDrive - Folkehelseinstituttet/VKM Data",
  "27.02.2025_maur_forprosjekt_biologisk_mangfold/data/species/1317433",
  "risk_assessment_anoplolepis_gracilipes_20251126_201333.html"
)

if (file.exists(gpt_file)) {
  htmltools::tags$iframe(
    src = gpt_file,
    width = "100%",
    height = "450px",
    style = "border: 1px solid #ccc;"
  )
} else {
  cat("GPT-Researcher report not found.")
}
```